{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-14 thesis/overview.tex
overview 1st p pronouns

\subsubsection{Simultaneity}
This simultaneity of threads is achieved either by executing each one on a separate CPU, or by interleaving them nondeterministically (as controlled by clock interrupts) on the same CPU.
Because clock interrupts can occur at any instruction\footnote{
	With some exceptions in kernel-level programming, which I discuss later.
},
we consider single-CPU multithreading to be simultaneous at the granularity of individual instructions.
%
Likewise, when multiple CPUs access the same memory,
hardware protocols generally ensure that the events of a single instruction are executed atomically from the perspective of all CPUs.
Although there are some exceptions --
unlocked memory-to-memory instructions,
unaligned writes \cite{unaligned-writes},
and weak memory consistency models \cite{memory-consistency-models} --
we model multicore concurrency the same way as above,
deferring these exceptions beyond the scope of this work.
We refer to an execution trace depicting the global sequence of events as a {\em thread interleaving} or {\em schedule}.

@@

Each such abstraction provides certain semantics about what thread interleavings can arise surrounding their use.
When building a tool for testing concurrent programs,
one may include some computational understanding of the behaviour of any, or all, of these abstractions.
Annotating a certain abstraction's semantics treats it as a trusted concurrency primitive in its own right,
and allows the testing tool to reduce the possible space of interleavings (or the set of false positive data-race candidates reported, etc.),
at the cost of increasing the implementation and theoretical complexity of the analysis.
In this thesis, I will consider locks and descheduling to be the only concurrency primitives,
and assume the others listed above are implemented using those as building blocks (an exercise for the reader \cite{thrlib}).

@@

Concerning possible execution patterns,
the main difference between STM and HTM is the circumstances under which a transaction may abort.
A software-backed transaction will abort if and only if a memory conflict occurs therein with another thread.
HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
I will explore the consequences of this difference further in Chapter~\ref{chap:tm}.

In this thesis, I will focus on HTM as my platform for testing transactional programs,
to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.

@@

Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
From the perspective of such tools, state spaces such as these wherein equivalent states may be reached by multiple paths
are represented as a directed acyclic graph (DAG) instead of as a tree.
This approach is called {\em stateful} model checking.
In this thesis, I focus on {\em stateless} model checking (and execution trees, not DAGs),
which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
Henceforth I will abbreviate ``stateless model checking'' simply as ``model checking'' for brevity.
Also, the term ``state space'' was originally coined to refer to the stateful approach's emphasis on the DAG's nodes
(i.e., program states);
while stateless checkers emphasize the tree's branches (i.e., execution sequences) instead,
I will continue to use ``state space'' for consistency with prior work.

@@

{\bf Concretely speaking:}
Figure~\ref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
The specifics of execution before the thread 1/thread 2 sequence,
other possible threads to run instead of threads 1 or 2,
and what logic the program executes in those subtrees
are all presumably arbitrary.
In these two highlighted branches,
if the transitions of threads 1 and 2 are {\em independent},
%if the operations performed by threads 1 and 2 are independent
%(i.e., no write/read or write/write access pairs to the same memory),
DPOR deduces that the subsequent program states (indicated by the red arrow) are equivalent.
Thence, only one of the two interleavings and its respective subtree needs to be executed
in order to check all possible program states.
I explain how DPOR implements such a deduction in more detail in \sect{\ref{sec:landslide-dpor}}.

@@

In this thesis I will tackle Pebbles and Pintos, two different system architectures used in educational operating systems courses.
This section describes the projects which students implement and which Landslide tests.

@@

The 15-410 course staff is notoriously secretive about the nature of many concurrency bugs
students commonly encounter during P2 and P3.
This is driven by a desire to cause students to find, diagnose, and fix these bugs on their own during the projects,
rather than to be surprised by them afterwards during grading
\cite{de0u-2018}.
%
One such example is the {\tt paraguay} unit test distributed with P2 (\sect{\ref{sec:education-pebbles-tests}}),
which targets a subtle condition-variable bug.
The test uses the {\tt misbehave} system call to target a particular thread interleaving likely to expose the bug
which is otherwise very unlikely to arise in normal execution.
The reference kernel specification \cite{kspec} does not define the {\tt misbehave} modes' behaviours,
as doing so would deprive students of the learning experience of discovering the interleaving in question on their own.
%
In this thesis I will occasionally use intentionally vague phrasing to preserve the mystery of these bugs.

@@

Some schools further offer a virtual memory project, extending the provided VM with a frame table and supplemental page table and fault handler \cite{standford-cs140,uchicago-cs230}, or supplemental HTTP server and {\tt malloc} assignments \cite{berkeley-cs162}.
Being largely architectural/algorithmic projects rather than concurrency-oriented ones, I am not concerned with these assignments in this thesis.
The main concurrency challenges in Pintos projects arise from the {\em threads} and {\em userprog} assignments:
implementing a correct {\tt alarm} routine,
ensuring the priority scheduler remains safe in the presence of concurrent threads of the same priority,
and designing correct interactions between the {\tt wait} and {\tt exit} system calls.

>>>>

\subsubsection{Simultaneity}
This simultaneity of threads is achieved either by executing each one on a separate CPU, or by interleaving them nondeterministically (as controlled by clock interrupts) on the same CPU.
Because clock interrupts can occur at any instruction\footnote{
	With some exceptions in kernel-level programming, discussed later.
},
we consider single-CPU multithreading to be simultaneous at the granularity of individual instructions.
%
Likewise, when multiple CPUs access the same memory,
hardware protocols generally ensure that the events of a single instruction are executed atomically from the perspective of all CPUs.
Although there are some exceptions --
unlocked memory-to-memory instructions,
unaligned writes \cite{unaligned-writes},
and weak memory consistency models \cite{memory-consistency-models} --
we model multicore concurrency the same way as above,
deferring these exceptions beyond the scope of this work.
We refer to an execution trace depicting the global sequence of events as a {\em thread interleaving} or {\em schedule}.

@@

Each such abstraction provides certain semantics about what thread interleavings can arise surrounding their use.
When building a tool for testing concurrent programs,
one may include some computational understanding of the behaviour of any, or all, of these abstractions.
Annotating a certain abstraction's semantics treats it as a trusted concurrency primitive in its own right,
and allows the testing tool to reduce the possible space of interleavings (or the set of false positive data-race candidates reported, etc.),
at the cost of increasing the implementation and theoretical complexity of the analysis.
This thesis will consider locks, descheduling, and transaction begin/end as the only concurrency primitives,
and assume the others listed above are implemented using those as building blocks (an exercise for the reader \cite{thrlib}).

@@

Concerning possible execution patterns,
the main difference between STM and HTM is the circumstances under which a transaction may abort.
A software-backed transaction will abort if and only if a memory conflict occurs therein with another thread.
HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
Chapter~\ref{chap:tm} will explore the consequences of this difference further.
%
This thesis will focus on HTM as my platform for testing transactional programs,
to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.

@@

Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
From the perspective of such tools, state spaces such as these wherein equivalent states may be reached by multiple paths
are represented as a directed acyclic graph (DAG) instead of as a tree.
This approach is called {\em stateful} model checking.
This thesis focuses on {\em stateless} model checking (and execution trees, not DAGs),
which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
Henceforth ``stateless model checking'' will be abbreviated simply as ``model checking'' for brevity.
Also, the term ``state space'' was originally coined to refer to the stateful approach's emphasis on the DAG's nodes
(i.e., program states);
while stateless checkers emphasize the tree's branches (i.e., execution sequences) instead,
I will continue to use ``state space'' for consistency with prior work.

@@

{\bf Concretely speaking:}
Figure~\ref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
The specifics of execution before the thread 1/thread 2 sequence,
other possible threads to run instead of threads 1 or 2,
and what logic the program executes in those subtrees
are all presumably arbitrary.
In these two highlighted branches,
if the transitions of threads 1 and 2 are {\em independent},
%if the operations performed by threads 1 and 2 are independent
%(i.e., no write/read or write/write access pairs to the same memory),
DPOR deduces that the subsequent program states (indicated by the red arrow) are equivalent.
Thence, only one of the two interleavings and its respective subtree needs to be executed
in order to check all possible program states.
\sect{\ref{sec:landslide-dpor}} explains how DPOR implements such a deduction in more detail.

@@

This thesis will tackle Pebbles and Pintos, two different system architectures used in educational operating systems courses.
This section describes the projects which students implement and which Landslide tests.

@@

The 15-410 course staff is notoriously secretive about the nature of many concurrency bugs
students commonly encounter during P2 and P3.
This is driven by a desire to cause students to find, diagnose, and fix these bugs on their own during the projects,
rather than to be surprised by them afterwards during grading
\cite{de0u-2018}.
%
One such example is the {\tt paraguay} unit test distributed with P2 (\sect{\ref{sec:education-pebbles-tests}}),
which targets a subtle condition-variable bug.
The test uses the {\tt misbehave} system call to target a particular thread interleaving likely to expose the bug
which is otherwise very unlikely to arise in normal execution.
The reference kernel specification \cite{kspec} does not define the {\tt misbehave} modes' behaviours,
as doing so would deprive students of the learning experience of discovering the interleaving in question on their own.
%
I will occasionally use intentionally vague phrasing to preserve the mystery of these bugs.

@@

Some schools further offer a virtual memory project, extending the provided VM with a frame table and supplemental page table and fault handler \cite{standford-cs140,uchicago-cs230}, or supplemental HTTP server and {\tt malloc} assignments \cite{berkeley-cs162}.
Being largely architectural/algorithmic projects rather than concurrency-oriented ones, this thesis is not concerned with these assignments.
The main concurrency challenges in Pintos projects arise from the {\em threads} and {\em userprog} assignments:
implementing a correct {\tt alarm} routine,
ensuring the priority scheduler remains safe in the presence of concurrent threads of the same priority,
and designing correct interactions between the {\tt wait} and {\tt exit} system calls.
